Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-08-11 Manuscript/MPESubmission/Revision/response.tex

% -----------------
% Comment 4
% -----------------
\item{\textcolor{blue}{\textbf{Introduction:} Needs stronger emphasis on what specifically is unique about this study compared to what the many other studies on missing data have done and found.
Clearly outlining the questions in this study that are left open from previous research is important because some of the findings in this study suggest that there may be more error associated with missing data than previous research has suggested.}}

We have increased our emphasis on novelty throughout the introduction by more clearly stating the difference between the present study and the literature and more clearly outlining the question of this study, as follows:

% 0(before this section)-Missing data can be a problem
% 1-And until now, not many studies

% NC: Line numbers? TG: I'll do all line numbering once I've incorporated all the edits to the manuscript.
\textit{Until now, few attempts have been made to study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework (i.e. using both molecular and morphological data; e.g. Wiens et al., 2005; Manos et al., 2007; Pattinson et al., 2014).
% 2-These studies have done X and Y to see if missing data's a problem
These studies have assessed the effect of missing data on topology by either (1) comparing a dataset with missing data to subsets without missing data (Wiens et al. 2005); or (2) removing both molecular and some morphological data from living taxa to create artificial fossils (Manos et al., 2007; Pattinson et al., 2014).
% 3-They show that it's not a problem
Both approaches have shown that missing data are not a major problem and should not be an obstacle to combining both living and fossil species in the same phylogenies.
% 4-But they've done it a certain way
However, these studies were limited by their empirical-only approach and the use of only living taxa (Wiens et al., 2005) or the scheme of generating missing data (i.e. using the patterns from the fossil record; Manos et al., 2007; Pattinson et al., 2014).
% NC: This is still a problem (it's ok up to 4). WHY is an empirical-only approach limiting? WHY is it limiting to generate missing data using fossil patterns? 
Besides, the studies using fossil data, mainly focus on the palaeontological aspect of the question (i.e. the effect of missing data in fossils in Total Evidence matrices; Manos et al., 2007; Pattinson et al. 2014).} lines @@@.
% NC: I don't like all these i.e.s here. Why not be specific in the text rather than throwing everything into brackets? If you're only citing manos and patterson then just say what they did.
% (next §)5-We're doing it this way

\textit{In this study, we propose a theoretical assessment of the effect of missing data in the Total Evidence method by removing living or fossil data, or all data for certain characters.
% NC: Could be more specific here and have a sentence that says. "This is an advance on previous studies because..."
We test the effect of missing data by measuring two crucial aspects of topology in both Maximum Likelihood and Bayesian phylogenies: (i) the conservation of clades and (ii) the displacement of wild-card taxa.} lines @@@
% NC: Again how is this an advance. I guess it's related to this sentence from below:
% (cf. Wiens et al., 2005; Pattinson et al., 2014 measuring respectively only the clades support or the number of shared clades).

% TG: I actually rewrote this bit as well:
% NC: line numbers?
\textit{We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the resulting tree topology.
We inferred the topology from the matrices using both Maximum Likelihood and Bayesian Inference methods and measured the differences in topology using two different topological distance metrics as proxies for clade conservation (based on the Robinson-Foulds distance; Robinson and Foulds, 1981) and for wild-card taxa placement (based on the Triplets distance; Critchlow et al., 1996).
Thus, in the present study, we propose a new approach to the missing-data question, focusing on both palaeontological and neontological data (rather than just palaeontological data), using only simulated data and measuring the effects of missing data on two aspects of topology (cf. Wiens et al., 2005; Pattinson et al., 2014 measuring respectively only the clades support or the number of shared clades).} lines @@@
% NC: This paragraph seems to be repeating what you've already said above but in slightly more detail. Might make more sense to combine this paragraph and the one above it?

@@

Although though we agree with the comments of both reviewers that knowing the difference between the true and best/missing-data trees is important, we do not think that these results warrant inclusion in the main text.
In these simulations, we had to make a choice between (1) simulating larger matrices but with fewer chains (i.e. less variance) or (2) smaller matrices with more variance.
We deliberately made the second choice because our main question was "what is the effect of our parameters on recovering topology".
To answer this question, we maintain that comparing the best tree to the missing-data trees is the best approach for two reasons:
\begin{enumerate}
\item{The true tree is always unknown:} as we state in the manuscript: "In real world situations, the ``true'' tree is never available" (line 242 in the first manuscript and @@@ in the revision).
\item{The true tree is just the starting point of the simulations:} we use the true tree only as a starting point for our simulations.
Once generated, the true tree is fed to a cascade of algorithms and software (i.e. seqGen\{phyclust\}, rTraitDisc\{ape\}, RAxML, MrBayes) described in section 2.1 and 2.3 to get the best tree and the missing-data trees.
Comparing the latter to the true tree will answer our question "what is the effect of our missing-data parameters and the cascade of algorithms and software on recovering the true topology?".
\end{enumerate}
Therefore we added a second justification to the main text (see below this response) and added these analysis (true vs. best/missing-data trees) to the supplementary for readers that might be interested by this specific problem.

We added the following sentence to the main text to clarify and link to the supplementary analysis:

\textit{We compare ``best'' trees to ``missing data'' trees but could also compare ``true'' trees to the ``missing data'' trees.
In practice, the difference between the ``best'' trees and the ``missing data'' trees represents the effect of our missing data parameters and of the phylogenetic methods used to infer the ``missing data'' trees.
The difference between the ``true'' and the ``missing data'' trees, however, represents the effect of our parameters used to generate the ``true'' tree and the algorithms used to generate the ``complete'' matrix as well as the effect of our missing data parameters and the phylogenetic methods used.
Because the main aim of this study is to look at the effect our missing data parameters on topological recovery, we chose to represent only the comparisons between the ``best'' trees and ``missing data'' trees.
The results of the comparisons of the ``true'' and the ``missing data'' tree are available in appendix B.
Note that this makes little difference to our overall results.} line 276-280 @@@.

@@

See also appendix B section "Differences between the ``true'' and the ``best'' trees." and reviewer 2's comment 1 below.

% -----------------
% Comment 7
% -----------------
\item{\textcolor{blue}{\textbf{Methodology - Starting tree:} The authors should at least include some analysis in which no starting tree is given, the max. likelihood 'best' tree is given, and / or the 'best' tree is given but with some perturbation.
I fear that one of the main results may be biased due to this one methodological detail - a strong prior on tree topology.
The justification given was 'to speed up the Bayesian analysis' which does not seem adequate, given these datasets are small.}}

We took on board this comment and agree with the reviewer that if using the ``true'' tree as a starting tree had an effect it would bias our results.
We therefore ran an extra analysis to test if that was the case and found that it had no major effect of the inferred tree topologies.
We performed this extra analysis on a subset of five trees with the three missing data parameters at 0\%, 10\%, 25\%, 50\% and 75\% for 20 chains. 
We compared the topology of these trees in Bayesian inference when using the ``true'' tree as a starting tree or without using a starting tree (i.e. using the default in MrBayes which is to use a different random tree for each MCMC chain).
We found no significant difference in topologies between the Bayesian inferences starting with the true tree or a random tree.
Therefore, we do not believe our results are merely the result of using a specific starting tree topology.
These results are available in the supplementary materials Appendix A (Figure A2 and Table A1) and we also added a note to the manuscript:

\textit{We used a fixed starting tree rather than a random starting tree (default MrBayes; Ronquist et al., 2012b) to speed up our Bayesian inferences.
Note that a starting tree is not a Bayesian prior on topology \textit{per se} and using a fixed starting tree did not significantly affect topology compared to using random starting trees (see Appendix A, section ``Effect of the starting tree on Bayesian inference'').} line 323-327 @@@.
See also Appendix A section "Effect of the starting tree on Bayesian inference".

@@

% -----------------
% Comment 8
% -----------------
\item{\textcolor{blue}{\textbf{Results - Bhattacharyya meaning:} Some descriptive statistics are given, but it would help to tell us if these differences are meaningful.
For example, on line 414 the Bhattacharrya Coefficients are given suggesting RF distances are lower for ML than BI trees at 0.69, 0.48, and 0.66 - but if this coefficient represents the probability of overlap between distributions then shouldn't the coefficients be $<$0.05 to say that one is lower than the other?
In the table, mean and median values of $<$0.01 and $>$0.85 are highlighted as being the most extreme.
Later the focus on Bhattacharyya Coefficients is 1 when distributions overlap completely and 0 when they do not - should this be the criterion?
Just make it clear early on so the readers knows what to think when they see these numbers. }}

We clarified the interpretation of the Bhattacharyya Coefficient throughout the manuscript.
We say that two distributions are significantly different when their probability of overlapping (the Bhattacharyya Coefficient) is $>$0.05 and that they are significantly similar when this probability is $>$0.95.
Even though other values (e.g. 0.85) are not conclusive, we still reported them to show the relative overlap of the distributions (i.e. two distributions with a coefficient of 0.85 are not significantly similar but are closer to each other than two distributions with a coefficient of 0.15; which are themselves not significantly different but still less close to each other than the previous ones).
We added the following statement to the main text and to the appendix B:

\textit{The Bhattacharyya Coefficient is the probability of overlap between two distributions bounded between 0 (no overlap) and 1 (full overlap; Bhattacharyya, 1943, see Appendix B for calculation details).} lines @@@

\textit{When the Bhattacharyya Coefficient between two distributions is $<$0.05, the distributions are significantly different.
When this coefficient is $>$0.95, the distributions are significantly similar.
Values between these two thresholds show the probability of overlap between the distributions but do not allow us to define the significance of the similarity or differences between distributions.} line 419-424 @@@.

@@

The difference between the Bayesian consensus and the Bayesian posterior trees can be due to two factors:
\begin{enumerate}[I]
\item the way we assess topological recovery is slightly different in each case: for the Bayesian consensus trees, we simply compared the missing-data consensus tree to the best consensus tree; for the Bayesian posterior trees however, we randomly compared 1000 missing-data Bayesian posterior trees to 1000 best Bayesian posterior trees.
Therefore, even if the comparison methodology is the same, the fact that the second methods draws random trees means we expect the topological recovery score to be blurred in comparison to other comparisons that do not involve random comparisons.
\item the Bayesian posterior trees are always resolved and thus, in the same way as the Maximum Likelihood tree, more likely to contain false-positive nodes. On the other hand, because the consensus tree is a strict majority consensus rule, only nodes with a a good support will be represented in the tree and all other nodes will be collapsed. This reduces the number of false-positive nodes (since an unresolved tree is always better than a wrongly resolved one) and increases clade conservation but also increases the displacement of wild-card taxa (since it is likely that some Triplets will just not exist in a poorly-resolved consensus tree).
\end{enumerate}
Both points were discussed in the original manuscript lines 566-573 (or lines 645-652 @@@ in the revised version).
We expanded the paragraph to:

\textit{Additionally, the Bayesian posterior trees performed more poorly than the Bayesian consensus tree (figure 4, table 1 and appendix C figure C5 and tables C5, C6 and C7).
This may be because the Bayesian posterior trees are always resolved and thus more likely to contain incorrectly resolved nodes (i.e. decreasing the Normalised Robinson-Foulds metric).
Conversely, the Bayesian consensus tree might not resolve nodes that are poorly supported and thus is more likely to contain only correctly resolved nodes (i.e. increasing the Normalised Robinson-Foulds metric).} lines 652-658. @@@

@@

% -----------------
% Comment 10
% -----------------
\item{\textcolor{blue}{\textbf{Results - Bayesian Triplets:} Another important result that is not mentioned is that using the wildcard taxon metric (Tr), Bayesian consensus trees do quite poorly with missing data.
Values in Figures 4 and 5 clearly drop below zero, although negative values are not illustrated on the axis, and this indicates that the tree topologies are no better than, or even worse than, random trees.
It is especially true when fossils are missing data, and that is going to be an important finding with practical application but must be addressed. }}

We modified the figures 4 and 5 to show Normalised Triplets metric values below 0 and added a dashed line at 0 to underline the difference in scale between both metrics.
As mentioned above (reviewer 1 comment 9), this is due to the fact that the Bayesian consensus tree will be less and less resolved when missing data increases.
This has two related effects.
The first one being to retain "good" clade conservation (i.e. avoiding false-positive nodes by collapsing unsupported ones).
And the second one is that when these clades are collapsed, it reduces the number of available triplets and thus increases the wild-card taxa score (Normalised Triplets metric).
In fact, when a clade ((a;b);c); is collapsed to (a;b;c), the whole clade is conserved (a, b and c are still in a common clade) however they also all become wild-card taxa: the position of a regarding b and c is now incorrect.

We added a note on this in the section about the effects of the tree inference method:

\textit{The Bayesian consensus trees, however, perform badly for the Normalised Triplets metric: some parameter combinations, especially when the $M_F$ parameter reaches 75\% missing data, lead to negative values (figure 5).
A Normalised Triplets metric value below 0 means that the placement of some taxa is worse than expected by just randomly placing this taxon in the tree.
This can be interpreted as the absence of comparable triplets between some of the ``missing data" trees and ``best'' trees.
Even if clades are conserved (figure 5), the resolution within them can be poor to non-existent when a lot of data are missing (i.e. 75\%).
In such cases, the fossil taxa are likely to be placed in any of the clades that share the most characters.
These results are in agreement with previous studies that have showed that missing data can cause problems for recovering ``correct'' topologies, especially for small matrices of 100 characters (Wiens 2003).
It is important to note, however, that this effect can be reduced by increasing the number of characters (Wiens 2003).} lines 632-644. @@@

@@

% -----------------
% Comment 11
% -----------------
\item{\textcolor{blue}{\textbf{Conclusion:} The authors should clarify how their results compare to those of previous studies more directly to emphasize the nuances that make this study unique. }}

We rewrote the conclusion to deal with this comment (and the specific comments 13, 52 - see comment 52 for details on the choice of citing Arcila et al 2015. - and 56), as follows:

% TG: I rephrased as follows.
\textit{Previous studies have explored the effect of missing morphological data in Total Evidence matrices (Wiens et al., 2005; Manos et al.,2007, Pattinson et al., 2014).
The conclusions of theses studies, however, were limited by their empirical approach as well as several aspects of their experimental design (e.g. not including fossil taxa, Wiens et al., 2005; or focusing on the palaeontological aspect, Manos et al.,2007, Pattinson et al., 2014).
% NC: Again needs a bit more explanation. WHY is empirical approach a problem? Because it only gives answers to one specific scenario or clade?
Here we used a more in depth approach where missing data were generated from simulated data and according to three clearly defined missing-data parameters ($M_{L}$, $M_{F}$ or $N_{C}$).
This allowed us to confirm previous results that missing data can be especially problematic in small matrices (Wiens 2003), but also revealed the crucial importance of coding morphological data for living species in Total Evidence phylogenies.
Missing data in Total Evidence matrices is not a problem for recovering the ``best'' tree topology as long as enough living and fossil taxa in the matrix have data for overlapping morphological characters.
When missing data increases in any of our missing data parameters ($M_{L}$, $M_{F}$ or $N_{C}$), it reduces support for the placement of fossil taxa and increases the displacement of wildcard taxa.
Therefore we advise increased focus on coding morphological characters for a large number of the living taxa present in the matrix (i.e. 50\%) if the goal is to accurately combine both living and fossil species in phylogenies.
Doing so will increase overlap of morphological characters among living and fossil taxa, allowing the fossil taxa to be positioned relative to the living taxa based on their shared derived characters rather than simply on available data.} lines @@@

\textit{Additionally, the topology of the Bayesian consensus trees, regardless of the amount of missing data, were always closer to the ``best'' tree topology than the Maximum Likelihood trees.
This has also been observed in empirical data (e.g. Arcila et al., 2015) where Maximum Likelihood trees inferred from a Total Evidence matrix were less supported than the Bayesian consensus tree.
This might have an important impact on estimating topologies in the Total Evidence framework, because previous studies had to rely either on molecular scaffolds (e.g. Slater, 2013), taxonomic constraints (e.g. Slater, 2013; Beck and Lee, 2014) or even on fixing the topology (e.g. Ronquist et al., 2012a).
Therefore, we suggest extracting such topological backbones from the Bayesian consensus tree if needed.} lines @@@
% TG: maybe add in a non tip-dating analysis way (i.e. Nick Matzke comment)? 
% NC: Sure if it fits. If you want to "cite" Nick you can use (N.Matzke pers.comm.)

\textit{To conclude, the results of our analyses are encouraging and show that it is possible to accurately combine both neontological and palaeontological morphological data in the same phylogeny as long as there is sufficient overlap.
Hopefully, using these approaches will greatly improve our understanding of macroevolutionary patterns and processes.} lines 602-634. @@@
% NC: I don't like this last bit. It's really weak. I'd cut it to be honest. The first sentence is not very clear (suddenly you revert back to neo and paleo) and the second sentence has no real explanation or link to the rest of the paper.

@@

% -----------------
% Comment 12
% -----------------
\item{\textcolor{blue}{\textbf{Figure 4:} Figure 4 has a color scheme issue, the colors in the caption do not match up to the figure.}}

We apologies for this silly mistake and thank the reviewer for pointing it out. We changed the colors to match with the paper's color scheme: \textit{Maximum Likelihood trees (black), Bayesian consensus trees (grey), Maximum Likelihood bootstrap trees (blue) and Bayesian posterior tree distributions (orange)}. See figure 4 caption.

@@

We changed the axes of the graphs showing Normalised Triplets metric to show values below 0 throughout the manuscript and the appendices.

% -----------------
% Comment 14
% -----------------
\item{\textcolor{blue}{\textbf{Table caption:} Table caption should include a line about 'pooling across missing data schemes' to clarify what is presented.}}

We clarified the caption of the Table 1 as follows:

\textit{Each line summarizes the probabilities of overlap between the distributions of the ``best'' tree versus trees from each inference method (Maximum Likelihood; Bayesian consensus; Maximum Likelihood Bootstraps and Bayesian posterior trees) pooled across all combinations of missing data parameter values, using the Normalised Robins-Foulds (RF) and Triplets (Tr) metrics. 
Values highlighted in bold are the extreme values of high or low probability of overlap between two methods. If two methods have a high probability of overlap, they have a similar ability to recover the ``correct'' tree topology."}
% NC: Do you want to say values > 0.95 are significantly similar, values < 0.05 are significantly different?
